Clustering Neural Populations by Poisson Dynamic Factor Model

### 2022-01-20 latexDraft/draft.tex
draft 1 update

	\begin{abstract}
	Modern recording techniques allow neuroscientists to study multiple neural populations over extended time periods in a large-scale, and the relationships within and between populations are summarized by low-dimensional latent vectors. When the neural activities are globally nonlinear, using single population analysis is inappropriate. However, defining the populations is usually difficult and  wrong cluster assignments will lead to bias in latent structure inferences. To tackle this challenge, we develop a clustering method based mixture of Poisson dynamic factor model. The number of cluster is treated as a parameter in mixture of finite mixtures (MFM) model, and the posteriors are sampled by a MCMC algorithm. To sample the posteriors efficiently, we approximate the full conditional distribution of latent state by Gaussian and approximate the marginal likelihood by making use of the Poisson-Gamma conjugacy. We further apply our method to neuropixel data and hippocampus data for illustration.
	\end{abstract}

@@

	The bias term ($\bm{b}^{(j)}$) in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$. Allowing the time-varying bias term will lead to infinite solutions of linear dynamics. If the $\bm{A}^{(j)}$ is one solution, then $\bm{A}^{*(j)} = \bm{A}^{(j)} + \bm{M}$ and $\bm{b}_t^{*(j)} = \bm{b}_t^{(j)} - \bm{M}\bm{x}^{(j)}_t$ will be another set of solution, for any $\bm{M} \in \mathbb{R}^{p\times p}$. The same reason for using time constant $g^{(j)}$.

@@

	\begin{align*}
		K &\sim p_k &&\text{where}\, p_k \, \text{is a p.m.f. on} \{1,2,\ldots\}\\
		\bm{\pi}=(\pi_1,\ldots,\pi_k) &\sim Dir_k(\gamma, \ldots,\gamma) &&\text{given} K=k\\
		Z_1,\ldots,Z_n&\stackrel{i.i.d.}{\sim}\pi &&\text{given} \pi\\
		\bm{\theta}_1,\ldots,\bm{\theta}_k&\stackrel{i.i.d.}{\sim}\bm{H} &&\text{given} K=k\\
		\bm{Y}_i = (y_{i1},\ldots,y_{iT})' &\sim SSFM(\bm{\theta}^{(z_i)}) &&\text{independently for}\, i=1,\ldots,N, \text{given}\, \bm{\theta}_{1:K}\, \text{and}\, Z_{1:N}
	\end{align*}

	Besides the conceptual correctness, using MFM model allows us integrate the prior knowledge easily. Moreover, compared to DPM, MFM has some better properties for clustering, for example, MFM posterior on number of cluster is more concentrated and consistent, and MFM tend to give clusters size at the same order of magnitude while DPM may lead to a few large clusters and many small clusters. See [reference] for detailed discussion.

@@

	In this paper, we choose to do inference by MCMC. Because of the Poisson likelihood, the latent state $\bm{X}^{(j)}$has no closed full conditional distribution. We can sample the posterior by particle MCMC directly, but this can be slow. However, due to the Markovian structure of the model, the conditional log-posterior is concave and its Hessian is block-tridiagonal. Thus, we can do the global Laplace approximation efficiently in $\mathcal{O}(T)$ \citep{Paninski2010}.The cluster index and number of cluster are sampled by the analog of partition-based algorithm in DPM \citep{Neal2000}. See details of the MCMC in appendix.

@@

	We first simulated three neural populations, with 10 neurons in each. We set $q=2$ for each cluster and the recording length is T=1000. After checking the trace plots up to 10,000 iterations, the convergence achieved after several steps. The panel A and B in figure 1 show the posterior mean firing rate and fitted latent state, averaging from iteration 500 to 1000. The results are transformed to PLDS parametrization for comparison. Notice that the PLDS parametrization also doesn't have the unique solution, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.\\
	\textbf{Still has the sign flipping issue. Fix later.}

	\begin{figure}[h!]
		\centering
		\subfigure[a]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
		\subfigure[b]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{image002.jpg}}%

		\subfigure[c]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
		\caption{Model Global Non-linearity by Clustering}
	\end{figure}

	Then we held out $1/4$ and $1/2$ data as test set in a “speckled” pattern, i.e. randomly select subset of data for each neuron as held-out dataset. Then we fit the model with and without clusters, keeping the same latent dimension, i.e. (1) 3 clusters with $p=1$ for each and (2) 1 cluster with $p=2\cdot 3 - 1 = 5$. The procedure is replicated for 100 times for two proportions, and the difference of held-out likelihood per spike between 2 fittings are shown in panel C in figure 1. The difference between (1) and (2) are always positive, and this shows doing the mixture of PDFM performs better than single PDFM. Moreover, as the proportion of training decrease (less data), the benefit of clustering becomes more significant. This suggests that doing clustering is necessary.

	\subsubsection{Clustering}

	Then use the same setting, we remove the label and use the mixture model to do the clustering by MFM. The prior of the cluster number is $K\sim Geometric(0.2)$. The panel A of figure 2 shows the trace of label $z_i$ for each neuron for the first 100 iterations.

	\begin{figure}[h!]
		\centering
		\subfigure[a]{\label{fig:2a}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
		\subfigure[b]{\label{fig:2b}\includegraphics[width=0.5\textwidth]{image005.jpg}}%

		\subfigure[c]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
		\subfigure[c]{\label{fig:2d}\includegraphics[width=0.5\textwidth]{image007.jpg}}%
		\caption{Clustering}
	\end{figure}

	In this toy simulation example, the signal is strong and the spiking amplitudes of neurons in each population are similar. This might be too simplified for real situation. In practice, the neural activity is usually sparse and can be recorded in high resolution. Moreover, the spiking amplitude for neuron in one population vary a lot. Therefore, we simulate another more realistic example. In this simulation, there are 3 clusters and the dimension of latent state is $p=1$ for each. However, there are 20 neurons with wider range but smaller value of loading $c_i$. The simulated firing rate is shown in panel B in figure 2. The trace plot of number of cluster (panel C in figure 2) shows that the model tend to further split some clusters into sub-population, based on the spiking amplitude. The similarity matrix (panel D in figure 2) of posterior, averaging from iteration 500 to 1000, shows that the model tend to split cluster 1 and 3 into two sub-clusters.

@@

	The Neuropixels dataset contains recording of neural activities in different brain regions, when doing the drift-grating experiment. Here, we only consider neurons with SNR > 3 and the spiking counts > 1000. Then, we use the recording activity from Lateral posterior nucleus of the thalamus (LP, 20 neurons), anteromedial visual area (VISam, 12 neurons) and ventral posteromedial nucleus of the thalamus (VPM, 14 neurons) during the spontaneous period. The bin size is 0.1s and truncate the data up to 1000 steps (100s recording) for convenience.

	Panel A in figure 3 shows the spiking counts of these 46 neurons. Here we first fit model using all data, with $p=1$ and $K\sim Geometric(0.3)$. For the number of clusters, panel B in figure 2 shows the trace plot for the first 5000 iterations and panel C shows the histogram of iteration 2500 to 5000. These plots show that these neurons are quite non-homogenous and they tend to form many sub-populations.The posterior similarity matrix, averaging from iteration 2500 to 5000 (panel D in figure 3), shows that there are several sub-populations in each anotomical cluster, and there are some tiny confusions between VISam and VPM.

	\textcolor{red}{\textbf{I tried to held-out half of the data in a speckled pattern and fit the model with clustering on and off (single population). The dimension is selected by held-out likelihood, which is p=1 for clustering on and p=2 for single population analysis. The trace plots of the held-out log-likelihood per spike (starting from iteration 2) shows that doing clustering doesn’t improve things… Maybe because “turning clustering on” introduce too much variance… The benefit of clustering will pop out if we use the deterministic algorithm, such as VB mentioned.}}

	\begin{figure}[h!]
		\centering
		\subfigure[a]{\label{fig:3a}\includegraphics[width=0.4\textwidth]{image008.jpg}}%
		\subfigure[b]{\label{fig:3b}\includegraphics[width=0.3\textwidth]{image009.jpg}}%

		\subfigure[c]{\label{fig:3c}\includegraphics[width=0.3\textwidth]{image010.jpg}}%
		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image011.jpg}}%

		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image012.jpg}}%
		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image013.jpg}}%

		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image001.png}}%
		\subfigure[c]{\label{fig:3d}\includegraphics[width=0.4\textwidth]{image002.png}}%
		\caption{Neuropixel}
	\end{figure}

@@

	\textcolor{red}{\textbf{What kind of story do I need to tell here? This should be different from the story from Neuropixel data. Maybe focus on the potential improvement of held-out likelihood after clustering? But I guess similar problem will happen. When fitting data with 84 neurons, the number of component jumps around 10.}}

@@

	\subsection{Full Conditional Distributions}

	\underline{\textit{$x_t^{(j)}$ and $\mu_t^{(j)}$}}:
	Let the $t^{\text{th}}$ column observation and firing rate of the $j^{\text{th}}$ cluster be $\widetilde{\bm{y}}_t = vec\left(\{y_{it}^{(j)}|z_i = j\}\right)$ and $\widetilde{\bm{\lambda}}_t^{(j)} = vec\left(\{\lambda_{it}|z_i = j\}\right)$. The number of neurons in cluster $j$ is $n_j=\left|\{i:z_i = j\}\right|$. The corresponding loading for these $n_j$ neurons is $\bm{C}^{(j)}\in\mathbb{R}^{n_j\times p}$, such that $\log \widetilde{\bm{\lambda}}_t^{(j)} = \mu_t^{(j)}\bm{1}_{n_j} + \bm{C}^{(j)}\bm{x}_t^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)\cdot \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$. Denote $\bm{\widetilde{C}}^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)$, $\bm{\widetilde{x}}_t^{(j)} = \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$, $\bm{\widetilde{x}}^{(j)} = \left(\bm{\widetilde{x}}_1^{'(j)},\ldots,\bm{\widetilde{x}}_T^{'(j)}\right)'$, $\widetilde{\bm{A}}^{(j)} = diag(f^{(j)}, \bm{A}^{(j)})$, $\widetilde{\bm{b}}^{(j)} = \left(g^{(j)}, \bm{b}^{'(j)}\right)'$ and $\widetilde{\bm{Q}}^{(j)} = diag(\sigma^{2(j)}, \bm{Q}^{(j)})$. The full conditional distribution $P(\bm{\widetilde{x}}^{(j)}|\ldots) = P(\bm{\widetilde{x}}^{(j)}|\{\widetilde{\bm{y}}_t\}_{t=1}^T, \bm{\widetilde{C}}^{(j)}, \widetilde{\bm{A}}^{(j)}, \widetilde{\bm{b}}^{(j)}, \widetilde{\bm{Q}}^{(j)})$ is approximated by a global Laplace approximation, i.e. $P(\bm{\widetilde{x}}^{(j)}|\ldots) \approx  N_{(p+1)T}(\bm{\widetilde{x}}^{(j)} | \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}, \bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}})$, with $\bm{\mu}_{\widetilde{\bm{x}}^{(j)}} = \argmax_{\widetilde{\bm{x}}^{(j)}}P(\bm{\widetilde{x}}^{(j)}|\ldots)$ and $\bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}}) = -\left(\nabla\nabla \log P(\bm{\widetilde{x}}^{(j)}|\ldots)\right |_{\bm{\widetilde{x}}^{(j)} = \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}} )^{-1}$. The log full conditional distribution $h(\bm{\widetilde{x}}^{(j)}) = \log P(\bm{\widetilde{x}}^{(j)}|\ldots)$ is given by:
	\begin{align*}
		h &= \text{const} + \sum_{t=1}^{T} \widetilde{\bm{y}}'_t\bm{\widetilde{C}}^{(j)}\bm{\widetilde{x}}_t^{(j)} - \widetilde{\bm{\lambda}}_t
		- \frac{1}{2}(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}_0(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})\\ 
		&-\sum_{t=2}^{T}\frac{1}{2}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})
	\end{align*}

	The $h(\bm{\widetilde{x}}^{(j)})$ is concave and hence unimodal, and the Markovian structure of the latent dynamics, and hence the tri-block-diagonal Hessian, makes it possible to compute a Newton update in $\mathcal{O}(T)$ \citep{Paninski2010}.

@@

	\underline{\textit{$\bm{c}_i$}}: 
	When writing the observation mapping in matrix form,
	$\log\bm{\lambda}_i = \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i$, the update of $\bm{c}_i$ reduces to a regular Poisson regression problem, given $\bm{\mu}^{(j)}$ and $\bm{X}^{(j)}$ are known. Here, we sample the posterior by no a No U-Turn Sampler (NUTS, \cite{Hoffman2011}).

>>>>

	\begin{abstract}
	Modern recording techniques allow neuroscientists to study multiple neural populations over extended time periods in a large-scale, and the relationships within and between populations are summarized by low-dimensional latent vectors. When the neural activities are globally nonlinear, using single population analysis is inappropriate. However, defining the populations is usually difficult and  wrong cluster assignments will lead to bias in latent structure inferences. To tackle this challenge, we develop a clustering method based mixture of Poisson dynamic factor model. The number of cluster is treated as a parameter in mixture of finite mixtures (MFM) model, and the posteriors are sampled by a MCMC algorithm. To sample the posteriors efficiently, we approximate the full conditional distribution of latent state by Gaussian and approximate the marginal likelihood by making use of the Poisson-Gamma conjugacy. We further apply our method to neuropixel data (and hippocampus data) for illustration.
	\end{abstract}

@@

	The bias term ($\bm{b}^{(j)}$) in the latent state is constant across time, unlike ind PLDS, which is  $\bm{b}_t^{(j)}$. When the bias term changes across the time, the update order of $\bm{A}^{(j)}$ and $\bm{b}_t^{(j)}$ will influence the results (usually, the $\bm{A}^{(j)}$ is updated first). When ignoring the updating order, the linear dynamics will have infinite solutions. If the $\bm{A}^{(j)}$ is one solution, then $\bm{A}^{*(j)} = \bm{A}^{(j)} + \bm{M}$ and $\bm{b}_t^{*(j)} = \bm{b}_t^{(j)} - \bm{M}\bm{x}^{(j)}_t$ will be another set of solution, for any $\bm{M} \in \mathbb{R}^{p\times p}$. This causes some inconvenience for inference. The same reason for using the time-constant $g^{(j)}$.

@@

	\begin{align*}
		K &\sim p_K &&\text{where}\, p_K \, \text{is a p.m.f. on} \{1,2,\ldots\}\\
		\bm{\pi}=(\pi_1,\ldots,\pi_k) &\sim Dir_k(\gamma, \ldots,\gamma) &&\text{given} K=k\\
		Z_1,\ldots,Z_n&\stackrel{i.i.d.}{\sim}\pi &&\text{given} \pi\\
		\bm{\theta}_1,\ldots,\bm{\theta}_k&\stackrel{i.i.d.}{\sim}\bm{H} &&\text{given} K=k\\
		\bm{Y}_i = (y_{i1},\ldots,y_{iT})' &\sim SSFM(\bm{\theta}^{(z_i)}) &&\text{independently for}\, i=1,\ldots,N, \text{given}\, \bm{\theta}_{1:K}\, \text{and}\, Z_{1:N}
	\end{align*}

	Besides the conceptual correctness, using MFM model allows us integrate the prior knowledge easily. Moreover, compared to DPM, MFM has some better properties for clustering, for example, MFM posterior on number of cluster is more concentrated and consistent, and MFM tend to give clusters size at the same order of magnitude while DPM may lead to a few large clusters and many small clusters. See \cite{Miller2018} for detailed discussion.

@@

	In this paper, we choose to do inference by MCMC. Because of the Poisson likelihood, the latent state $\bm{X}^{(j)}$has no closed full conditional distribution. We can sample the posterior by particle MCMC directly, but this can be slow. However, due to the Markovian structure of the model, the conditional log-posterior is concave and its Hessian is block-tridiagonal. Thus, we can do the global Laplace approximation efficiently in $\mathcal{O}(T)$ \citep{Paninski2010}.The cluster index and number of cluster are sampled by the analog of partition-based algorithm in DPM \citep{Neal2000}. See details of the MCMC update in the Appendix.

@@

	We first simulated three neural populations, with 10 neurons in each. We set $q=2$ for each cluster and the recording length is T=1000. The $\delta_i = 1$ for $i=1,\ldots, 10$ and is $0$ for the remaining neurons. The $\bm{G}^{(j)}$ is $0.1(1,\ldots,10)'\otimes\bm{1}'_2 + \bm{1}_{10}\otimes(-1, 2)$ to set up a gradient for the loading in each cluster and to differ the contribution of two latent vectors. We allow $\bm{A}^{(j)}$ be non-diagonal and possible interactions between clusters, which is used in single population PLDS model. The diagonal element is generated from $N(0.92, 0.1^2)$. The bias terms are $\bm{b}^{(j)}$ is $10^{-3}\bm{1}_2$ for the first 2 clusters and $10^{-4}\bm{1}_2$ for the last cluster. The $\bm{Q}^{(j)}$ are $10^{-3}\bm{I}_2$, $10^{-4}\bm{I}_2$ and $10^{-5}\bm{I}_2$.

	The Figure \ref{fig:1a} shows the trace plot for the log-likelihood per spike, up to 10,000 iterations. The plot shows (Figure \ref{fig:1b}) that the convergence achieved after several steps. Therefore, we show the results by averaging iteration 500 to 1000. Figure \ref{fig:1c} shows the posterior mean of the mean firing rate, while Figure \ref{fig:1d} shows the fitted latent state. The latent vectors are transformed to PLDS parametrization for comparison. Notice that the PLDS parametrization also doesn't have the unique solution, we let $\bm{G}^{(j)}$ has mean zeros columns and is orthogonal.

	Then we held out $1/4$ and $1/2$ data as test set in a “speckled” pattern, i.e. randomly select subset of data for each neuron as held-out dataset. The "speckled" cross-validation was previously used in \cite{Williams2020}. We then fit the model with and without clusters, keeping the same latent dimension, i.e. (1) 3 clusters with $p=1$ for each and (2) 1 cluster with $p=2\cdot 3 - 1 = 5$. The procedure is replicated for 100 times for two proportions, and the difference of held-out likelihood per spike between 2 fittings are shown in Figure \ref{fig:1e}. The difference between (1) and (2) are always positive, and this shows doing the mixture of PDFM performs better than single PDFM. Moreover, as the proportion of training decrease (less data), the benefit of clustering becomes more significant. This suggests that doing clustering is necessary.

	\begin{figure}[h!]
		\centering
		\subfigure[Trace plot of  log-likelihood per spike]{\label{fig:1a}\includegraphics[width=0.5\textwidth]{2_llhd1.png}}%
		\subfigure[Trace plot of  log-likelihood per spike, truncated (1000)]{\label{fig:1b}\includegraphics[width=0.5\textwidth]{3_llhd2.png}}%

		\subfigure[true vs. fitted firing rate]{\label{fig:1c}\includegraphics[width=0.5\textwidth]{image001.jpg}}%
		\subfigure[true vs. fitted latent vectors]{\label{fig:1d}\includegraphics[width=0.5\textwidth]{LF.png}}%

		\subfigure[cross-validation: clustered vs. single population]{\label{fig:1e}\includegraphics[width=0.5\textwidth]{image003.jpg}}%
		\subfigure[trace plot for non-labeled case]{\label{fig:1f}\includegraphics[width=0.5\textwidth]{image004.jpg}}%
		\caption{Simple case}
	\end{figure}

	\subsubsection{Clustering}
	Then use the same setting, we remove the label and use the mixture model to do the clustering by MFM. The prior of the cluster number is $K\sim Geometric(0.2)$. Figure \ref{fig:1f} shows the trace of label $z_i$ for each neuron for the first 100 iterations.

	\begin{figure}[h!]
		\centering
		\subfigure[Firing rate]{\label{fig:2a}\includegraphics[width=0.4\textwidth]{image005.jpg}}%
		\subfigure[trace plot for cluster number]{\label{fig:2b}\includegraphics[width=0.4\textwidth]{image006.jpg}}%
		\subfigure[similarity matrix]{\label{fig:2c}\includegraphics[width=0.4\textwidth]{image007.jpg}}%
		\caption{More complicated case: more variety of amplitude \& weaker signal}
	\end{figure}

	In this toy simulation example, the signal is strong and the spiking amplitudes of neurons in each population are similar. This might be too simplified for real situation. In practice, the neural activity is usually sparse and can be recorded in high resolution. Moreover, the spiking amplitudes for neuron in one nominated population vary a lot, and hence may form a few subpopulations. Therefore, we then simulate another more realistic example. In this simulation, there are 3 clusters and the dimension of latent state is $p=1$ for each. However, there are 20 neurons with wider range but smaller value of loading $c_i$, corresponding to potential subpopulations and weak signal. The simulated firing rate is shown in Figure \ref{fig:2a}. The trace plot of number of cluster (Figure \ref{fig:2b}) shows that the model tend to further split some clusters into sub-population, based on the spiking amplitude. This is more clearly shown in the posterior similarity matrix ((Figure \ref{fig:2c}), averaging from iteration 500 to 1000. In similarity matrix, the entry $(i,j)$ is the posterior probability that data points $i$ and $j$ belong to the same cluster. The rows and columns are ordered according to the ground truth. The similarity matrix shows that the model tend to split cluster 1 and 3 into two sub-clusters.

@@

	The Neuropixels dataset contains recording of neural activities in different brain regions of mouse. Here, we only consider neurons with SNR > 3 and the spiking counts > 1000. Then, we first use the recording activity from Lateral posterior nucleus of the thalamus (LP, 20 neurons), anteromedial visual area (VISam, 12 neurons) and ventral posteromedial nucleus of the thalamus (VPM, 14 neurons) during the spontaneous period. The bin size is 0.1s and truncate the data up to 1000 steps (100s recording) for convenience.

	Figure \ref{fig:3a} shows the spiking counts of these 46 neurons. Here we first fit model using all data, with $p=1$ and $K\sim Geometric(0.3)$. For the number of clusters, Figure \ref{fig:3b} shows the trace plot for the first 5000 iterations. All the following results are from averaging samples from iteration 2500 to 5000. The Figure \ref{fig:3c} shows the histogram of the posterior samples. These plots show that these neurons are quite non-homogeneous and they tend to form many sub-populations. These 3 nominated populations by anatomy tend to form around 10 (sub)-clusters.

	The Figure \ref{fig:3d} and Figure \ref{fig:3e} show the trace of log-likelihood per spike. The fitted mean firing rate is shown in Figure \ref{fig:3f}. When investigating the similarity matrix in Figure \ref{fig:3g} (sorted according to the anatomical population, such that 1-20 are LP, 21-32 are VISam and 33-46 are VPM), there are several subpopulations in each anatomical cluster. Generally, the clustering results are consistent with the anatomical assignment, although there are some tiny confusions between VISam and VPM.

	However, this is not the case when the mouse is in the experiment. We choose the same set of neurons but use recordings when the mouse are watching drift gratings. The Figure \ref{fig:3h} shows the trace plot of cluster number. The Figure \ref{fig:3i} and Figure \ref{fig:3j} show the fitted firing rate and similarity matrix, again averaging from iteration 2500 to 5000.

	When comparing the observed spiking activities, it's fairly hard to see differences when the mouse is exposed to the stimuli and not. However, the posterior similarity matrix in the experiment shows that there's more confusion between LP and VISam, compared to the spontaneous response. This suggests that the functional connection pattern between neurons in different areas when the stimuli are changed.

	\textcolor{red}{\textbf{I tried to held-out half of the data in a speckled pattern and fit the model with clustering on and off (single population). The dimension is selected by held-out likelihood, which is p=1 for clustering on and p=2 for single population analysis. The trace plots of the held-out log-likelihood per spike (starting from iteration 2) shows that doing clustering doesn’t improve things (although they are very close)… Maybe because “turning clustering on” introduce too much variance… The benefit of clustering will pop out if we use the deterministic algorithm, such as VB mentioned.}}

	\begin{figure}[h!]
		\centering
		\subfigure[spontaneous: spike counts]{\label{fig:3a}\includegraphics[width=0.5\textwidth]{image008.jpg}}%
		\subfigure[spontaneous: cluster number trace]{\label{fig:3b}\includegraphics[width=0.35\textwidth]{image009.jpg}}%
		\subfigure[spontaneous: cluster number histogram]{\label{fig:3c}\includegraphics[width=0.35\textwidth]{image010.jpg}}%

		\subfigure[spontaneous: llhd/spk trace]{\label{fig:3d}\includegraphics[width=0.3\textwidth]{image001.png}}%
		\subfigure[spontaneous: llhd/spk trace, truncated]{\label{fig:3e}\includegraphics[width=0.3\textwidth]{image002.png}}%
		\subfigure[spontaneous: true vs. fitted firing rate]{\label{fig:3f}\includegraphics[width=0.3\textwidth]{image011.jpg}}%
		\subfigure[spontaneous: similarity matrix]{\label{fig:3g}\includegraphics[width=0.3\textwidth]{image012.jpg}}%

		\subfigure[drift-grating: cluster number trace]{\label{fig:3h}\includegraphics[width=0.4\textwidth]{12_numTrace.png}}%
		\subfigure[drift-grating: true vs. fitted firing rate]{\label{fig:3i}\includegraphics[width=0.4\textwidth]{16_fitFR.png}}%
		\subfigure[drift-grating: similarity matrix]{\label{fig:3j}\includegraphics[width=0.4\textwidth]{14_simPixel.png}}%
		\caption{Neuropixel}
	\end{figure}

@@

	\textcolor{red}{\textbf{I prefer just give one application.} Because 1) there's a page limit (9 pages excluding ref \& appendix in 2021, and now it seems we will exceed the limit), and 2) the V1 data is enough to illustrate the idea If we really need to add the second example, we should tell the story in another angle}.

@@

	\subsection{MCMC updates}
	The posteriors are sampled by a Gibbs sampler, the full conditional distributions, if available, are shown as follows:

	\underline{\textit{Update $x_t^{(j)}$ and $\mu_t^{(j)}$}}:
	Let the $t^{\text{th}}$ column observation and firing rate of the $j^{\text{th}}$ cluster be $\widetilde{\bm{y}}_t = vec\left(\{y_{it}^{(j)}|z_i = j\}\right)$ and $\widetilde{\bm{\lambda}}_t^{(j)} = vec\left(\{\lambda_{it}|z_i = j\}\right)$. The number of neurons in cluster $j$ is $n_j=\left|\{i:z_i = j\}\right|$. The corresponding loading for these $n_j$ neurons is $\bm{C}^{(j)}\in\mathbb{R}^{n_j\times p}$, such that $\log \widetilde{\bm{\lambda}}_t^{(j)} = \mu_t^{(j)}\bm{1}_{n_j} + \bm{C}^{(j)}\bm{x}_t^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)\cdot \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$. Denote $\bm{\widetilde{C}}^{(j)} = \left(\bm{1}_{n_j}, \bm{C}^{(j)} \right)$, $\bm{\widetilde{x}}_t^{(j)} = \left(\mu_t^{(j)}, \bm{x}_t^{'(j)}\right)'$, $\bm{\widetilde{x}}^{(j)} = \left(\bm{\widetilde{x}}_1^{'(j)},\ldots,\bm{\widetilde{x}}_T^{'(j)}\right)'$, $\widetilde{\bm{A}}^{(j)} = diag(f^{(j)}, \bm{A}^{(j)})$, $\widetilde{\bm{b}}^{(j)} = \left(g^{(j)}, \bm{b}^{'(j)}\right)'$ and $\widetilde{\bm{Q}}^{(j)} = diag(\sigma^{2(j)}, \bm{Q}^{(j)})$. The full conditional distribution $P(\bm{\widetilde{x}}^{(j)}|\ldots) = P(\bm{\widetilde{x}}^{(j)}|\{\widetilde{\bm{y}}_t\}_{t=1}^T, \bm{\widetilde{C}}^{(j)}, \widetilde{\bm{A}}^{(j)}, \widetilde{\bm{b}}^{(j)}, \widetilde{\bm{Q}}^{(j)})$ is approximated by a global Laplace approximation, i.e. $P(\bm{\widetilde{x}}^{(j)}|\ldots) \approx  N_{(p+1)T}(\bm{\widetilde{x}}^{(j)} | \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}, \bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}})$, with $\bm{\mu}_{\widetilde{\bm{x}}^{(j)}} = \argmax_{\widetilde{\bm{x}}^{(j)}}P(\bm{\widetilde{x}}^{(j)}|\ldots)$ and $\bm{\Sigma}_{\widetilde{\bm{x}}^{(j)}}) = -\left(\nabla\nabla \log P(\bm{\widetilde{x}}^{(j)}|\ldots)\right |_{\bm{\widetilde{x}}^{(j)} = \bm{\mu}_{\widetilde{\bm{x}}^{(j)}}} )^{-1}$. The log full conditional distribution $h(\bm{\widetilde{x}}^{(j)}) = \log P(\bm{\widetilde{x}}^{(j)}|\ldots)$ is given by:
	\begin{align*}
		h &= \text{const} + \sum_{t=1}^{T} \left(\widetilde{\bm{y}}'_t\bm{\widetilde{C}}^{(j)}\bm{\widetilde{x}}_t^{(j)} - \widetilde{\bm{\lambda}}_t\right)
		- \frac{1}{2}(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}_0(\bm{\widetilde{x}}_1^{(j)}  - \bm{\widetilde{x}}_0^{(j)})\\ 
		&-\sum_{t=2}^{T}\frac{1}{2}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})'\bm{\widetilde{\bm{Q}}}^{(j)}(\bm{\widetilde{x}}_t^{(j)}  - \widetilde{\bm{A}}^{(j)}\bm{\widetilde{x}}_{t-1}^{(j)} - \widetilde{\bm{b}}^{(j)})
	\end{align*}
	, where "const" represents the constant. The $h(\bm{\widetilde{x}}^{(j)})$ is concave and hence unimodal, and the Markovian structure of the latent dynamics, and hence the tri-block-diagonal Hessian, makes it possible to compute a Newton update in $\mathcal{O}(T)$ \citep{Paninski2010}. The same technique is used in the E-step for PLDS model \citep{Macke2011}.

@@

	\underline{\textit{Update $\bm{c}_i$}}: 
	When writing the observation mapping in matrix form,
	$\log\bm{\lambda}_i = \bm{\mu}^{(j)}+ \bm{X}^{(j)}\bm{c}_i$, the update of $\bm{c}_i$ reduces to a regular Poisson regression problem, given $\bm{\mu}^{(j)}$ and $\bm{X}^{(j)}$ are known. Here, we sample the posterior by no a No U-Turn Sampler (NUTS, \cite{Hoffman2011}), within the Gibbs sampler.

	\underline{\textit{Update Linear dynamics of latent state}}: The parameters for linear dynamics are $f^{(j)}$, $g^{(j)}$, $\sigma^{2(j)}$, $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$. To make the model identifiable, we simply assume $\bm{A}^{(j)} = diag(a_1^{(j)},\ldots, a_p^{(j)})$ and $\bm{Q}^{(j)} = diag(q_1^{(j)},\ldots, q_p^{(j)})$. Therefore, we can update $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$ dimension-by-dimension, as the update in $f^{(j)}$, $g^{(j)}$ and $\sigma^{2(j)}$. Use the priors $\sigma^{2(j)} \sim IG\left(\nu_0/2, \nu_0\sigma_0^2/2\right)$ and  $\left( g^{(j)}, f^{(j)}\right)'\sim N(\bm{\mu}_0, \sigma^{2(j)}\bm{\Lambda}_0^{-1})$. The prior $\mu_0$ is specified as $(0,1)'$. Denote $\bm{y}_{\mu^{(j)}} = \left(\mu_2^{(j)}, \ldots, \mu_T^{(j)}\right)'$ and $\bm{X}_{\mu^{(j)}} = \left(\bm{1}_{T-1}, \bm{\mu}_{1:(T-1)}^{(j)}\right)$, with $\bm{\mu}_{1:(T-1)}^{(j)} = \left(\mu_1^{(j)}, \ldots, \mu_{T-1}^{(j)}\right)'$. The full conditional distributions are $\sigma^{2(j)}|\ldots \sim IG\left(\frac{\nu_0 + T-1}{2}, \frac{\nu_0\sigma^2_0 + \bm{y}'_{\mu^{(j)}}\bm{y}_{\mu^{(j)}} + \bm{\mu}'_0\bm{\Lambda}_0\bm{\mu}_0 - \bm{\mu}'_n\bm{\Lambda}_n\bm{\mu}_n }{2}\right)$ and $\left( g^{(j)}, f^{(j)}\right)'|\ldots \sim N(\bm{\mu}_n, \sigma^{2(j)}\bm{\Lambda}_n^{-1})$, with $\bm{\Lambda}_n = \bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}} + \bm{\Lambda}_0$ and $\bm{\mu}_n = \bm{\Lambda}_n^{-1}\left(\bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}}\left(\bm{X}'_{\mu^{(j)}}\bm{X}_{\mu^{(j)}}\right)^{-1}\bm{X}'_{\mu^{(j)}}\bm{y}_{\mu^{(j)}} + \bm{\Lambda}_0\bm{\mu}_0\right)$. The updates of $\bm{A}^{(j)}$, $\bm{b}^{(j)}$ and $\bm{Q}^{(j)}$ are similar, with diagonal assumption.
